---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# betanorm

<!-- badges: start -->
<!-- badges: end -->

**betanorm** is a small toolkit of fast C++ array-normalization utilities
([Rcpp](https://cran.r-project.org/package=Rcpp) /
[RcppArmadillo](https://cran.r-project.org/package=RcppArmadillo)):

- **Quantile normalization** — map each sample row onto a target distribution
  (no missing values; a focused port of
  `preprocessCore::normalize.quantiles.use.target()`).
- **BMIQ-style gold-standard calibration** — for DNA methylation beta matrices
  when a pipeline expects beta-mixture quantile calibration to a reference
  profile. The default path (`nL = 3`, `niter = 5`) is intentionally
  legacy-compatible, not a claim that five outer EM steps are statistical
  convergence.

## Installation

Install the development version from [GitHub](https://github.com/hhp94/betanorm):

``` r
# install.packages("pak")
pak::pak("hhp94/betanorm")
```

## Quantile normalization

`quantile_norm()` maps each matrix row onto a target vector by rank: equal
length uses Bolstad-style indexing; unequal length uses linear quantile
interpolation. Inputs must be finite with no missing values.

```{r setup}
library(betanorm)
```

```{r quantile-norm}
set.seed(1)
X <- matrix(rnorm(5 * 100), nrow = 5)
target <- sort(rnorm(100))
Y <- quantile_norm(X, target)
dim(Y)
```

## Gold-standard calibration (BMIQ-style)

For methylation betas, `bmiq_calibration()` fits a two- or three-component
beta mixture per sample and maps components toward a gold-standard profile.
Use this when you need that procedure (e.g. clock-style workflows); prefer
`quantile_norm()` when a plain target distribution is enough.

Built-in example data: gold-standard annotation and a small beta matrix
(6 samples × ~21k probes).

```{r bmiq-data}
data(horvath_goldstandard)
data(GPL21145_sample)

dim(GPL21145_sample)
head(horvath_goldstandard$goldstandard2, 3)
```

### Legacy defaults (`nL = 3`, `niter = 5`)

Three-state BMIQ with a hard outer EM cap of **five** iterations is the
intentional default. That pair reproduces common legacy implementations
(Horvath-style / MEAT-style clocks and similar pipelines). **Do not raise
`niter` on the `nL = 3` path** unless you deliberately want results that
diverge from those implementations—`bmiq_calibration()` warns if you do.

If you want the mixture EM to run longer or toward convergence, switch to
**`nL = 2`** (U/M only; no intermediate H step) and raise `niter` / tighten
`fit.policy` there instead.

```{r calibrate}
fit <- bmiq_calibration(
  datM = GPL21145_sample,
  goldstandard.beta = horvath_goldstandard$goldstandard2,
  verbose = FALSE
)

fit
calibrated <- as.matrix(fit)
dim(calibrated)
```

### Two-component path (`nL = 2`)

Use `nL = 2` for unmethylated / methylated only (`doH` defaults to `FALSE`)
and whenever you prefer longer EM rather than the legacy five-iteration
three-state cap.

```{r calibrate-nl2}
fit2 <- bmiq_calibration(
  datM = GPL21145_sample,
  goldstandard.beta = horvath_goldstandard$goldstandard2,
  nL = 2L,
  niter = 25L,
  verbose = FALSE
)
fit2
```

### Sample-level error handling

By default a failed sample stops the call. To keep going and record failures:

```{r error-policy, eval = FALSE}
fit <- bmiq_calibration(
  datM = datM,
  goldstandard.beta = gold,
  on.sample.error = "continue",
  failed.sample = "NA", # or "original"
  h.policy = "optional", # keep U/M if H fails
  fit.policy = "usable" # accept finite (not fully converged) fits
)
```

## Main functions

| Function | Role |
|----------|------|
| `quantile_norm()` | Quantile-normalize matrix rows to a target distribution |
| `bmiq_calibration()` | BMIQ-style calibration to a gold-standard beta profile |
| `horvath_goldstandard` | Gold-standard probe annotation + mean betas |
| `GPL21145_sample` | Example beta matrix for demos and tests |

## References

Bolstad BM, Irizarry RA, Åstrand M, Speed TP (2003). A comparison of
normalization methods for high density oligonucleotide array data based on
variance and bias. *Bioinformatics* 19(2), 185–193.
<https://doi.org/10.1093/bioinformatics/19.2.185>

Teschendorff AE, Marabita F, Lechner M, Bartlett T, Tegner J, Gomez-Cabrero D,
Beck S (2013). A beta-mixture quantile normalization method for correcting
probe design bias in Illumina Infinium 450 k DNA methylation data.
*Bioinformatics* 29(2), 189–196.
<https://doi.org/10.1093/bioinformatics/bts680>
